---
title: "Cohort Utilization & Cost Enrichment"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cohort Utilization & Cost Enrichment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE,
  warning = FALSE,
  message = FALSE
)
```

# Cohort Utilization & Cost Enrichment

In Health Economics and Outcomes Research (HEOR) and Real-World Evidence (RWE), evaluating Healthcare Resource Utilization (HCRU) and direct medical costs requires capturing granular patient encounters across multiple care settings (inpatient, outpatient, emergency, pharmacy, diagnostics, procedures, and financial claims).

Following the **DARWIN EU / OHDSI** package ecosystem standards (`CohortConstructor`, `PatientProfiles`, `CohortCharacteristics`), **HERMES** provides a pipe-friendly, 3-layer architecture for cohort construction, in-database column enrichment, and standardized reporting.

---

## 3-Layer Architecture Overview

```text
┌────────────────────────────────────────────────────────────────────────────────────────┐
│                                   OMOP CDM DATABASE                                    │
│  (visit_occurrence, provider, drug_exposure, procedure_occurrence, measurement, cost)  │
└────────────────────────────────────────────────────────────────────────────────────────┘
                                    │
       ┌────────────────────────────┼────────────────────────────┐
       ▼                            ▼                            ▼
 1. Episode Constructors     2. Cohort Enrichers          3. Summarisation & Reporting
 (CohortConstructor Style)   (PatientProfiles Style)      (CohortCharacteristics Style)
 ─────────────────────────   ───────────────────────      ────────────────────────────
 Standalone episode cohorts  Appends windowed metric      Standardised result objects
 in write schema:            columns in-database:         for publication tables:
 • computeHospitalization*   • addInpatients()            • summariseUtilization()
 • computeInfusionCohorts()  • addEmergencyCare()         • summariseCosts()
                             • addOutpatientVisits()      • tableUtilization()
                             • addPrescriptions()         • tableCosts()
                             • addProcedures()            • plotUtilization()
                             • addCosts()                 • plotCosts()
                             • addVisits() (composite)
```

---

## 1. Setup & CDM Connection

We initialize an OMOP CDM reference (such as the mock HERMES test database):

```{r setup}
library(hermes)
library(dplyr)
library(omopgenerics)

# Connect to OMOP CDM
cdm <- mockHERMES()
```

---

## 2. Layer 1: Care Episode Cohort Constructors (`CohortConstructor` Style)

Care episode constructors collapse contiguous or overlapping raw encounters into discrete hospitalization or infusion therapy episodes and register them in the database write schema.

### Hospitalization & Readmission Episodes

```{r episodes_hosp}
# Construct collapsed hospitalization episodes and 30-day readmissions
cdm$hosp_episodes <- computeHospitalizationCohorts(
  cdm = cdm,
  name = "hosp_episodes",
  visitConceptIds = c(9201L, 8717L, 581379L),
  icuConceptIds = 32037L,
  readmissionWindow = 30L
)

# Inspect cohort entries
cdm$hosp_episodes |>
  dplyr::collect()
```

### Infusion Administration Episodes

```{r episodes_infusion}
# Construct parenteral/intravenous infusion therapy episodes
cdm$infusion_episodes <- computeInfusionCohorts(
  cdm = cdm,
  name = "infusion_episodes",
  routeConceptIds = c(4171047L) # Intravenous
)

cdm$infusion_episodes |>
  dplyr::collect()
```

---

## 3. Layer 2: In-Database Cohort Enrichment (`PatientProfiles` Style)

Cohort enricher functions take an existing study cohort, execute database-side joins across configurable temporal windows (e.g., baseline `[-365, -1]`, 1-year follow-up `[0, 365]`, or full follow-up `[0, Inf]`), and append metric columns without dropping zero-utilization subjects.

### Modular Enricher Pipeline

```{r enrich_pipeline}
# Pipe study cohort through all domain enrichers with fixed and open-ended windows
cdm$study_enriched <- cdm$target_cohort |>
  addInpatients(
    window = list(baseline = c(-365, -1), followup = c(0, 365), all_followup = c(0, Inf)),
    readmissions = TRUE,
    stratifySpecialty = TRUE,
    specialties = list(cardiology = 38004453L)
  ) |>
  addEmergencyCare(
    window = list(baseline = c(-365, -1), followup = c(0, 365), all_followup = c(0, Inf))
  ) |>
  addOutpatientVisits(
    window = list(baseline = c(-365, -1), followup = c(0, 365), all_followup = c(0, Inf)),
    stratifySpecialty = TRUE,
    includeEmergency = FALSE
  ) |>
  addPrescriptions(
    window = list(baseline = c(-365, -1), followup = c(0, 365), all_followup = c(0, Inf)),
    daysSupply = TRUE,
    pdc = TRUE
  ) |>
  addProcedures(
    window = list(baseline = c(-365, -1), followup = c(0, 365), all_followup = c(0, Inf))
  ) |>
  addCosts(
    window = list(baseline = c(-365, -1), followup = c(0, 365), all_followup = c(0, Inf)),
    costField = "total_paid",
    name = "study_enriched"
  )

# Inspect enriched columns
colnames(cdm$study_enriched)
```

### All-In-One Multi-Setting Visits with `addVisits()`

To extract Inpatient, Outpatient, and Emergency utilization in a single step with unified specialty stratification:

```{r add_visits_demo}
cdm$visits_enriched <- cdm$target_cohort |>
  addVisits(
    window = list(baseline = c(-365, -1), all_followup = c(0, Inf)),
    settings = c("inpatient", "outpatient", "emergency"),
    stratifySpecialty = TRUE,
    specialties = list(
      oncology = c(38004507L, 38004006L),
      cardiology = 38004453L
    ),
    readmissions = TRUE,
    name = "visits_enriched"
  )

colnames(cdm$visits_enriched)
```

---

## 4. Layer 3: Analytics & Reporting (`CohortCharacteristics` Style)

We aggregate patient-level utilization and costs into DARWIN EU standardized `summarised_result` objects compatible with `visOmopResults`.

### Summarise Care Utilization

```{r summarise_util}
utilSummary <- summariseUtilization(
  cohort = cdm$study_enriched,
  strata = list("cohort_definition_id"),
  estimates = c("mean", "sd", "median", "q25", "q75", "min", "max")
)

# View standardised result structure
utilSummary |>
  dplyr::select("strata_name", "strata_level", "variable_name", "estimate_name", "estimate_value") |>
  utils::head(10)
```

### Summarise Direct Medical Costs

```{r summarise_cost}
costSummary <- summariseCosts(
  cohort = cdm$study_enriched,
  strata = list("cohort_definition_id"),
  estimates = c("mean", "sd", "median", "q25", "q75", "min", "max")
)

costSummary |>
  dplyr::select("strata_name", "strata_level", "variable_name", "estimate_name", "estimate_value") |>
  utils::head(8)
```

### Publication-Ready GT & Flextable Output

`tableUtilization()` and `tableCosts()` integrate with `visOmopResults::visOmopTable()` to format publication-grade summary tables:

```{r render_table}
# Format summary table with concise estimate strings
tableUtilization(
  result = utilSummary,
  estimateName = c(
    "Mean (SD)" = "<mean> (<sd>)",
    "Median [Q25 - Q75]" = "<median> [<q25> - <q75>]"
  ),
  header = c("cohort_name")
)
```

---

## 5. Visualizing Care Utilization & Costs (`visOmopResults` Style)

HERMES provides `plotUtilization()` and `plotCosts()` for rapid exploratory and publication visualization, leveraging standard `visOmopResults` ggplot themes and facets.

### Bar Plots for Care Utilization

```{r plot_util_bar, fig.width = 8, fig.height = 3.5}
plotUtilization(
  result = utilSummary,
  metric = "inpatient_admissions",
  plotType = "barplot"
)
```

### Box Plots for Direct Medical Costs

```{r plot_cost_box, fig.width = 8, fig.height = 4}
plotCosts(
  result = costSummary,
  costColumn = "cost_total",
  plotType = "boxplot"
)
```

---

## 6. Summary Matrix Mapping

| Domain | Metrics Extracted | Function Responsible | Output Columns |
| :--- | :--- | :--- | :--- |
| **[Inpatient]** | Admissions, ICU, LOS, Readmissions, Specialties | `addInpatients()` (`addHospitalizations()`) | `inpatient_admissions_*`, `inpatient_los_days_*`, `icu_admissions_*`, `readmissions_30d_*`, `{specialty}_inpatient_admissions_*` |
| **[Emergency]** | ER Visits & Specialist Acute Care | `addEmergencyCare()` (`addEmergency()`) | `emergency_visits_*`, `{specialty}_emergency_visits_*` |
| **[Outpatient]** | GP, Specialist, Other Visits, Specialties | `addOutpatientVisits()` | `gp_visits_*`, `specialist_visits_*`, `other_outpatient_visits_*`, `{specialty}_visits_*` |
| **[Multi-Setting]** | Inpatient + Outpatient + Emergency Visits | `addVisits()` | Combined Inpatient, Outpatient, Emergency columns |
| **[Pharmacy]** | Rx Fills, Days Supply, PDC, Infusions | `addPrescriptions()` | `rx_fills_*`, `days_supply_*`, `pdc_*`, `infusions_*` |
| **[Diagnostics/Proc]** | Labs, Imaging, Procedures | `addProcedures()` | `lab_tests_count_*`, `procedures_count_*` |
| **[Direct Costs]** | Inpatient, Outpatient, Drug, Total | `addCosts()` | `cost_inpatient_*`, `cost_outpatient_*`, `cost_drug_*`, `cost_total_*` |
