---
title: "Cohort Utilization & Cost Enrichment"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cohort Utilization & Cost Enrichment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE,
  warning = FALSE,
  message = FALSE
)
```

# Cohort Utilization & Cost Enrichment

In Health Economics and Outcomes Research (HEOR) and Real-World Evidence (RWE), evaluating Healthcare Resource Utilization (HCRU) and direct medical costs requires capturing granular patient encounters across multiple care settings (inpatient, outpatient, pharmacy, diagnostics, procedures, and financial claims).

Following the **DARWIN EU / OHDSI** package ecosystem standards (`CohortConstructor`, `PatientProfiles`, `CohortCharacteristics`), **HERMES** provides a pipe-friendly, 3-layer architecture for cohort construction, in-database column enrichment, and standardized reporting.

---

## 3-Layer Architecture Overview

```text
┌────────────────────────────────────────────────────────────────────────────────────────┐
│                                   OMOP CDM DATABASE                                    │
│  (visit_occurrence, provider, drug_exposure, procedure_occurrence, measurement, cost)  │
└────────────────────────────────────────────────────────────────────────────────────────┘
                                    │
       ┌────────────────────────────┼────────────────────────────┐
       ▼                            ▼                            ▼
 1. Episode Constructors     2. Cohort Enrichers          3. Summarisation & Reporting
 (CohortConstructor Style)   (PatientProfiles Style)      (CohortCharacteristics Style)
 ─────────────────────────   ───────────────────────      ────────────────────────────
 Standalone episode cohorts  Appends windowed metric      Standardised result objects
 in write schema:            columns in-database:         for publication tables:
 • computeHospitalization*   • addHospitalizations()      • summariseUtilization()
 • computeInfusionCohorts()  • addOutpatientVisits()      • summariseCosts()
                             • addPrescriptions()         • tableUtilization()
                             • addProcedures()            • tableCosts()
                             • addCosts()
```

---

## 1. Setup & CDM Connection

We initialize an OMOP CDM reference (such as the mock HERMES test database):

```{r setup}
library(HERMES)
library(dplyr)
library(omopgenerics)

# Connect to OMOP CDM
cdm <- mockHERMES()
```

---

## 2. Layer 1: Care Episode Cohort Constructors (`CohortConstructor` Style)

Care episode constructors collapse contiguous or overlapping raw encounters into discrete hospitalization or infusion therapy episodes and register them in the database write schema.

### Hospitalization & Readmission Episodes

```{r episodes_hosp}
# Construct collapsed hospitalization episodes and 30-day readmissions
cdm$hosp_episodes <- computeHospitalizationCohorts(
  cdm = cdm,
  name = "hosp_episodes",
  visitConceptIds = c(9201L, 8717L, 581379L),
  icuConceptIds = 32037L,
  readmissionWindow = 30L
)

# Inspect cohort entries
cdm$hosp_episodes |>
  dplyr::collect()
```

### Infusion Administration Episodes

```{r episodes_infusion}
# Construct parenteral/intravenous infusion therapy episodes
cdm$infusion_episodes <- computeInfusionCohorts(
  cdm = cdm,
  name = "infusion_episodes",
  routeConceptIds = c(4171047L) # Intravenous
)

cdm$infusion_episodes |>
  dplyr::collect()
```

---

## 3. Layer 2: In-Database Cohort Enrichment (`PatientProfiles` Style)

Cohort enricher functions take an existing study cohort, execute database-side joins across configurable temporal windows (e.g., baseline `[-365, -1]` and follow-up `[0, 365]`), and append metric columns without dropping zero-utilization subjects.

```{r enrich_pipeline}
# Pipe study cohort through all 5 domain enrichers
cdm$study_enriched <- cdm$target_cohort |>
  addHospitalizations(
    window = list(baseline = c(-365, -1), followup = c(0, 365)),
    readmissions = TRUE
  ) |>
  addOutpatientVisits(
    window = list(baseline = c(-365, -1), followup = c(0, 365)),
    stratifySpecialty = TRUE
  ) |>
  addPrescriptions(
    window = list(baseline = c(-365, -1), followup = c(0, 365)),
    daysSupply = TRUE,
    pdc = TRUE
  ) |>
  addProcedures(
    window = list(baseline = c(-365, -1), followup = c(0, 365))
  ) |>
  addCosts(
    window = list(baseline = c(-365, -1), followup = c(0, 365)),
    costField = "total_paid",
    name = "study_enriched"
  )

# Inspect enriched columns
colnames(cdm$study_enriched)
```

```text
The enriched cohort table contains:
• Inpatient: inpatient_admissions_*, inpatient_los_days_*, icu_admissions_*, icu_los_days_*, readmissions_30d_*
• Outpatient: emergency_visits_*, gp_visits_*, specialist_visits_*, other_outpatient_visits_*
• Pharmacy: rx_fills_*, days_supply_*, pdc_*, infusions_*
• Diagnostics: lab_tests_count_*, procedures_count_*
• Direct Costs: cost_inpatient_*, cost_outpatient_*, cost_drug_*, cost_procedure_*, cost_total_*
```

---

## 4. Layer 3: Analytics & Reporting (`CohortCharacteristics` Style)

We aggregate patient-level utilization and costs into DARWIN EU standardized `summarised_result` objects compatible with `visOmopResults`.

### Summarise Care Utilization

```{r summarise_util}
utilSummary <- summariseUtilization(
  cohort = cdm$study_enriched,
  strata = list("cohort_definition_id"),
  estimates = c("mean", "sd", "median", "q25", "q75", "min", "max")
)

# View standardised result structure
utilSummary |>
  dplyr::select("strata_name", "strata_level", "variable_name", "estimate_name", "estimate_value") |>
  utils::head(10)
```

### Summarise Direct Medical Costs

```{r summarise_cost}
costSummary <- summariseCosts(
  cohort = cdm$study_enriched,
  strata = list("cohort_definition_id"),
  estimates = c("mean", "sd", "median", "q25", "q75")
)

costSummary |>
  dplyr::select("strata_name", "strata_level", "variable_name", "estimate_name", "estimate_value") |>
  utils::head(8)
```

### Publication-Ready GT Table

```{r render_table}
# Render formatted table
tableUtilization(utilSummary, type = "tibble")
```

---

## 5. Summary Matrix Mapping

| Domain | Metrics Extracted | Function Responsible | Output Columns |
| :--- | :--- | :--- | :--- |
| **[Inpatient]** | Admissions, ICU, LOS, Readmissions | `addHospitalizations()` | `inpatient_admissions_*`, `inpatient_los_days_*`, `icu_admissions_*`, `readmissions_30d_*` |
| **[Outpatient]** | GP, Specialist, ER, Other Visits | `addOutpatientVisits()` | `gp_visits_*`, `specialist_visits_*`, `emergency_visits_*` |
| **[Pharmacy]** | Rx Fills, Days Supply, PDC, Infusions | `addPrescriptions()` | `rx_fills_*`, `days_supply_*`, `pdc_*`, `infusions_*` |
| **[Diagnostics/Proc]** | Labs, Imaging, Procedures | `addProcedures()` | `lab_tests_count_*`, `procedures_count_*` |
| **[Direct Costs]** | Inpatient, Outpatient, Drug, Total | `addCosts()` | `cost_inpatient_*`, `cost_outpatient_*`, `cost_drug_*`, `cost_total_*` |
